%************************************************
\chapter{Methodology}\label{ch:methodology}
%************************************************

This research was conducted at Neo4j, within the Graph Data Science team. The goal
was to build and evaluate a \acsfont{GNN} training pipeline that uses Neo4j as the
primary graph store.

The first step was to implement a naive \acsfont{GNN} training system coupled with a
graph database. This was compared with a PyTorch Geometric in-memory implementation
to verify that the graph-database backend did not degrade model quality, \ie that it
did not cause slower convergence or a worse final test accuracy. Training throughput
was not compared between the two implementations at this stage, since the PyG baseline
keeps the entire graph in \acsfont{RAM} while the graph-database implementation stores
the majority of the graph on disk.


%------------------------------------------------
\section{System Architecture}\label{sec:sys_arch}
%------------------------------------------------

The system consists of three main components: a Neo4j graph database acting as the
graph and feature store, a Python training process that implements the PyG abstract
interfaces, and a \texttt{NodeLoader} that orchestrates mini-batch construction during
training. The implementations \texttt{FeatureStore} and \texttt{GraphStore} acts as the interface
between the training process and the \texttt{Neo4j} database. The implementation of \texttt{BaseSampler}
defines how the subgraph sampling is being done.


%------------------------------------------------
\section{Graph Representation in Neo4j}\label{sec:graph_repr}
%------------------------------------------------

Nodes in Neo4j are stored with a label corresponding to the node type (\eg,
\texttt{Paper} for citation networks). Each node carries the following properties:

\begin{itemize}
    \item A unique integer id.
    \item Pre-computed node feature vectors. The vectors can either be stored as Float64[] or bytes[]
    \item A target class label for supervised node classification, which is an integer signifying what class the node belongs to
    \item A split indicator (\texttt{train}, \texttt{val}, \texttt{test}) used to
    partition the dataset. This can either be a string, or integer.
\end{itemize}

Edges are stored as typed relationships connecting pairs of nodes. No edge features
are used in the experiments.


%------------------------------------------------
\section{PyG Interface Implementations}\label{sec:pyg_impl}
%------------------------------------------------

We implement the three PyG abstract classes described in \autoref{sec:pyg}:\\

\noindent \textbf{\texttt{Neo4jAbstractFS}} is an abstract class extending FeatureStore. The
 purpose of this class is to provide storage of the feature vectors (e.g. a graph database), and
 to implement an efficient interface for fetching feature vectors. The reason it is kept abstract
is that while there are some methods any Neo4j interacting PyG FeatureStore needs (e.g. executing queries)
there are also some features that may vary depending on the use case. One example of this is caching.
Avoiding the neo4j fetching by storing commonly used features in RAM is a good optimization, but how this 
caching is done in the feature store is something that should be up to the user. There are also two extensions of the 
\texttt{Neo4jAbstractFS}, which in the \texttt{Neo4jCachedFS} and \texttt{Neo4jNoCacheFS}.\\

\noindent \textbf{\texttt{Neo4jAbstractGS}} an abstract class extending GraphStore, with the main goal
of implementing efficient traversal over the graph to enable fast subgraph sampling. It has two implementations
\texttt{Neo4jMultiGS} and \texttt{Neo4jSingleGS} where the difference is how many sampler they are 
optimized of being able to handle.\\

\noindent \textbf{\texttt{Neo4jSampler}} implements the sampling algorithm, and is tightly coupled
with the graph store where it executes the sampling algorithm.

%------------------------------------------------
\section{Sampling Algorithms}\label{sec:sampling_algs}
%------------------------------------------------

Two sampling strategies are evaluated:

\paragraph{Neighbor sampling.} For each seed node in the mini-batch, the sampler
issues a Cypher query that traverses the graph up to $K$ hops, sampling at most
$[k_0, k_1, \ldots, k_{K-1}]$ neighbors at each hop. Neighbors exceeding the limit
are selected uniformly at random without replacement. This corresponds to the GraphSAGE
sampling strategy.

\paragraph{GraphSAINT random-walk sampling.} Instead of expanding neighborhoods
around seed nodes, GraphSAINT constructs a subgraph by first sampling a set of nodes
via random walks and then inducing the subgraph over those nodes.


%------------------------------------------------
\section{Graph Neural Network Model}\label{sec:model}
%------------------------------------------------

All experiments use a two-layer Graph Convolutional Network (\acsfont{GCN}). The
model applies two \texttt{GCNConv} message-passing layers with \acsfont{ReLU}
activations and a final linear classifier. The architecture is intentionally kept
simple to focus evaluation on the data pipeline rather than the model itself.


%------------------------------------------------
\section{Datasets and Experimental Setup}\label{sec:datasets}
%------------------------------------------------

Experiments are conducted on the following datasets:

All experiments are run on a single machine with the Neo4j database instance running
locally. Hyperparameters such as learning rate, number of epochs, batch size, and
sampling limits are kept consistent across compared configurations and recorded in a
central configuration file.


%------------------------------------------------
\section{Benchmarking Methodology}\label{sec:benchmarking}
%------------------------------------------------

A dedicated \texttt{Measurer} class wraps the training loop and collects timing,
profiling, and resource utilization data for each run, by logging events to a csv file. the 
csv file is then compiled into plots, and other result files summarizing the experiment outcome


%------------------------------------------------
\subsection{End-to-End Latency Decomposition}\label{sec:e2e_latency}
%------------------------------------------------

Per-batch wall-clock time is decomposed into the following segments:

\begin{itemize}
    \item \textbf{DB server latency (topology / features).} The total time from query
    submission to consumption of the last result row ($t_{\text{last}}$), as reported
    by the Neo4j Bolt protocol summary. This includes query planning, execution,
    serialization, and network round-trip time. The time to the first result row
    ($t_{\text{first}}$) is indicated as a darker sub-segment within this bar.

    \item \textbf{Network + driver receive.} The additional client-side time spent
    receiving and deserializing Bolt protocol packets after the first row arrives,
    computed as client wall time minus $t_{\text{last}}$.

    \item \textbf{Python ETL.} Time spent constructing PyG data structures from raw
    query results (index remapping, tensor assembly).

    \item \textbf{GNN compute.} Time for the forward pass, loss computation, and
    backward pass.
\end{itemize}


%------------------------------------------------
\subsection{Query Execution Profiling}\label{sec:query_profiling}
%------------------------------------------------

Cypher queries are instrumented with the \texttt{PROFILE} keyword to collect
operator-level statistics: rows produced, \acsfont{DB} hits, and estimated memory
usage. These are aggregated across batches and stored as both averaged \acsfont{CSV}
summaries and raw plan trees (\texttt{query\_profile\_raw.json}) for offline
inspection.


%------------------------------------------------
\subsection{Python-Level Profiling}\label{sec:python_profiling}
%------------------------------------------------

Call-stack profiles are collected using Python's \texttt{cProfile} module and stored
as \texttt{train\_profile.txt}. These are used to identify hot-spots in the Python
data pipeline that are not visible in the query-level metrics, such as tensor
allocation overhead or unnecessary data copies.


%------------------------------------------------
\subsection{Sampler Comparison Framework}\label{sec:sampler_comparison}
%------------------------------------------------

A dedicated experiment runner executes multiple samplers under identical conditions
and aggregates their results for side-by-side comparison. Plots comparing accuracy
convergence, per-batch latency, and subgraph statistics are generated automatically
at the end of the experiment.


%------------------------------------------------
\section{Correctness and Verification}\label{sec:correctness}
%------------------------------------------------

The graph-database implementation is verified against the in-memory PyG baseline by
comparing final test accuracy and convergence curves across multiple runs. Subgraph
statistics (number of sampled nodes and edges per batch) are also checked to confirm
that the samplers produce subgraphs of the expected size distribution.
